---
title: "CnH2O2 growkar workflow example"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "dd-growkar-workflow-files/figure-gfm/"
)

if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  pkg_root <- "."
} else {
  candidates <- c(".", "..", "../..")
  match_idx <- which(file.exists(file.path(candidates, "DESCRIPTION")))
  pkg_root <- if (length(match_idx) > 0L) candidates[[match_idx[[1]]]] else NULL
}

if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)
```

## Prepare data

This example reads `CnH2O2_OD.txt`, converts it to the canonical tidy format
used by `growkar`, and validates the result.

```{r prepare-data}
dd_path <- if (file.exists("CnH2O2_OD.txt")) {
  "CnH2O2_OD.txt"
} else {
  system.file("extdata", "CnH2O2_OD.txt", package = "growkar")
}

dd <- read.delim(dd_path, check.names = FALSE)
tidy_dd <- as_tidy_growth_data(dd)
validate_growth_data(tidy_dd)

h2o2_levels <- tidy_dd |>
  dplyr::distinct(condition) |>
  dplyr::mutate(
    concentration_mM = as.numeric(sub("^H2O2\\(([-0-9.]+)mM\\)$", "\\1", .data$condition))
  ) |>
  dplyr::arrange(.data$concentration_mM) |>
  dplyr::pull(condition)

tidy_dd <- dplyr::mutate(
  tidy_dd,
  condition = factor(.data$condition, levels = h2o2_levels)
)

averaged_dd <- tidy_dd |>
  dplyr::group_by(.data$condition, .data$time) |>
  dplyr::summarise(od = mean(.data$od, na.rm = TRUE), .groups = "drop") |>
  dplyr::rename(sample = .data$condition)

head(tidy_dd)
```

## Plot growth curves with averaged replicates

This plot uses the averaged replicate trajectories for each condition and
returns a `ggplot2` object that can be customized further if needed.

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  averaged_dd,
  average_replicates = FALSE,
  colour_col = "sample",
  palette_name = "Dark2"
)
```

## Summarize doubling time with H2O2(0mM) as the reference

This summary compares replicate-level doubling times for each condition against
`H2O2(0mM)`.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  tidy_dd,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "H2O2(0mM)"
)

dt_stats <- dplyr::arrange(dt_stats, .data$condition)

knitr::kable(dt_stats, digits = 3)
```

## Plot doubling time comparisons

This plot shows mean doubling time with error bars and comparison brackets
against `H2O2(0mM)`.

```{r doubling-time-plot, fig.width=8, fig.height=5}
plot_doubling_time(
  tidy_dd,
  comparison_col = "condition",
  compare_to = "H2O2(0mM)",
  palette_name = "Dark2"
)
```

## Detect exponential phase in all averaged samples

This section inspects the highest-ranked candidate exponential windows across
all averaged H2O2 conditions.

```{r detect-exponential-phase}
phase_tbl <- dplyr::bind_rows(
  lapply(split(averaged_dd, averaged_dd$sample), detect_exponential_phase)
)

phase_tbl <- phase_tbl |>
  dplyr::group_by(.data$sample) |>
  dplyr::slice_head(n = 3) |>
  dplyr::ungroup()

knitr::kable(phase_tbl, digits = 3)
```

## Fit and plot one representative growth curve

This section fits a logistic model to the averaged `H2O2(0mM)` curve and
overlays observed and fitted values.

```{r fit-one-sample}
fit <- averaged_dd |>
  filter(sample == "H2O2(0mM)") |>
  fit_growth_curve(model = "logistic")

extract_params(fit)
```

```{r plot-fit, fig.width=8, fig.height=5}
plot_fitted_curve(fit)
```
