---
title: "CnH2O2 growkar workflow example"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "dd-growkar-workflow-files/figure-gfm/"
)

if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  pkg_root <- "."
} else {
  candidates <- c(".", "..", "../..")
  match_idx <- which(file.exists(file.path(candidates, "DESCRIPTION")))
  pkg_root <- if (length(match_idx) > 0L) candidates[[match_idx[[1]]]] else NULL
}

if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)
```

## Prepare data

This example reads `CnH2O2_OD.txt`, converts it to the canonical tidy format
used by `growkar`, and validates the result.

```{r prepare-data}
dd_path <- if (file.exists("CnH2O2_OD.txt")) {
  "CnH2O2_OD.txt"
} else {
  system.file("extdata", "CnH2O2_OD.txt", package = "growkar")
}

dd <- read.delim(dd_path, check.names = FALSE)
tidy_dd <- as_tidy_growth_data(dd)
validate_growth_data(tidy_dd)

h2o2_levels <- tidy_dd |>
  dplyr::distinct(condition) |>
  dplyr::mutate(
    concentration_mM = as.numeric(sub("^H2O2\\(([-0-9.]+)mM\\)$", "\\1", .data$condition))
  ) |>
  dplyr::arrange(.data$concentration_mM) |>
  dplyr::pull(condition)

tidy_dd <- dplyr::mutate(
  tidy_dd,
  condition = factor(.data$condition, levels = h2o2_levels)
)

head(tidy_dd)
```

## Coerce to SummarizedExperiment

`growkar` can also package the processed data into a lightweight
`growkar_data` object and coerce it into a `SummarizedExperiment` for
Bioconductor-oriented workflows.

```{r summarized-experiment}
growkar_obj <- as_growkar(tidy_dd)
se <- methods::as(growkar_obj, "SummarizedExperiment")
se
```

## Store growth metrics in SummarizedExperiment metadata

The same rule-based doubling-time summary can be computed directly on the
`SummarizedExperiment` object and stored in `metadata(se)$growth_metrics`.

```{r summarized-experiment-metrics}
se <- growth_metrics(
  se,
  method = "rule_based",
  comparison_col = "condition",
  compare_to = "H2O2(0mM)"
)

se_metrics <- S4Vectors::metadata(se)$growth_metrics |>
  dplyr::arrange(.data$condition)

knitr::kable(se_metrics, digits = 3)
```

## Plot growth curves with averaged replicates

This plot uses the averaged replicate trajectories for each condition and
returns a `ggplot2` object that can be customized further if needed.

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  tidy_dd,
  average_replicates = TRUE,
  colour_col = "condition",
  palette_name = "Dark2"
)
```

## Summarize doubling time with H2O2(0mM) as the reference

This summary compares replicate-level doubling times for each condition against
`H2O2(0mM)` using the rule-based exponential interval.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  tidy_dd,
  method = "rule_based",
  comparison_col = "condition",
  compare_to = "H2O2(0mM)"
)

dt_stats <- dplyr::arrange(dt_stats, .data$condition)

knitr::kable(dt_stats, digits = 3)
```

## Plot doubling time comparisons

This plot shows mean doubling time with error bars and comparison brackets
against `H2O2(0mM)` using the rule-based exponential interval.

```{r doubling-time-plot, fig.width=8, fig.height=5}
plot_doubling_time(
  tidy_dd,
  comparison_col = "condition",
  compare_to = "H2O2(0mM)",
  method = "rule_based",
  palette_name = "Dark2"
)
```
