---
title: "KN99 CDK7 growkar workflow example"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "dd-growkar-workflow-files/figure-gfm/"
)

if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  pkg_root <- "."
} else {
  candidates <- c(".", "..", "../..")
  match_idx <- which(file.exists(file.path(candidates, "DESCRIPTION")))
  pkg_root <- if (length(match_idx) > 0L) candidates[[match_idx[[1]]]] else NULL
}

if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)
```

## Import and validate data

This example reads `dose_response_BS181_20Dec24_Cdk7Tag.txt`, converts it to
the canonical tidy format used by `growkar`, and validates the result.

```{r prepare-data}
dd_path <- if (file.exists("dose_response_BS181_20Dec24_Cdk7Tag.txt")) {
  "dose_response_BS181_20Dec24_Cdk7Tag.txt"
} else {
  system.file("extdata", "dose_response_BS181_20Dec24_Cdk7Tag.txt", package = "growkar")
}

dd <- read.delim(dd_path, check.names = FALSE)
tidy_dd <- as_tidy_growth_data(dd)
validate_growth_data(tidy_dd)

head(tidy_dd)
```

## Create the canonical SummarizedExperiment

`growkar` can also package the processed data into a lightweight
`growkar_data` object and coerce it into a `SummarizedExperiment` for
Bioconductor-oriented workflows. This example keeps the full CDK7 dataset so
the facet plot can show `KN99`, `CM2444`, `CM2446`, and `CM2448`.

```{r summarized-experiment}
growkar_obj <- as_growkar(tidy_dd)
se <- methods::as(growkar_obj, "SummarizedExperiment")

se
```

## Analyze using SummarizedExperiment metadata

The same doubling-time summary can be computed directly on the
`SummarizedExperiment` object and stored in `metadata(se)$growth_metrics`.

```{r summarized-experiment-metrics}
se <- growth_metrics(
  se,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "KN99(0)"
)

se_metrics <- S4Vectors::metadata(se)$growth_metrics

knitr::kable(se_metrics, digits = 3)
```

## Plot growth curves with averaged replicates

This plot uses the averaged replicate trajectories for each condition and
returns a `ggplot2` object that can be customized further if needed.

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  se,
  average_replicates = TRUE,
  colour_col = "condition"
)
```

## Plot growth curves as facets

For complex datasets with multiple conditions such as `KN99(...)` and
`CM2448(...)`, `plot_growth_curve_facets()` averages replicates first and then
creates one facet per sample family such as `KN99`, `CM2444`, `CM2446`, and
`CM2448`.

```{r sample-facet-growth-curve, fig.width=8, fig.height=6}
plot_growth_curve_facets(
  se
)
```

## Summarize doubling time with KN99(0) as the reference

This summary compares replicate-level doubling times for each condition against
`KN99(0)` using the rolling-window exponential interval.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  se,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "KN99(0)"
)

knitr::kable(dt_stats, digits = 3)
```

## Plot doubling time comparisons

This plot shows mean doubling time with error bars and comparison brackets
against `KN99(0)` using the rolling-window exponential interval.

```{r doubling-time-plot, fig.width=8, fig.height=5}
plot_doubling_time(
  se,
  comparison_col = "condition",
  compare_to = "KN99(0)",
  method = "rolling_window"
)
```
