---
title: "growkar workflow for dd"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(growkar)
library(dplyr)
library(knitr)
```

## Prepare data

This example reads `CnH2O2_OD.txt`, converts it to the canonical tidy format
used by `growkar`, and validates the result.

```{r prepare-data}
dd_path <- if (file.exists("CnH2O2_OD.txt")) {
  "CnH2O2_OD.txt"
} else {
  system.file("extdata", "CnH2O2_OD.txt", package = "growkar")
}

dd <- read.delim(dd_path, check.names = FALSE)
tidy_dd <- as_tidy_growth_data(dd)
validate_growth_data(tidy_dd)

head(tidy_dd)
```

## Plot growth curves with averaged replicates

This plot averages replicates within each condition and returns a `ggplot2`
object that can be customized further if needed.

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  tidy_dd,
  average_replicates = TRUE,
  colour_col = "condition",
  palette_name = "Dark2"
)
```

## Summarize doubling time with H2O2(0mM) as the reference

This summary compares replicate-level doubling times for each condition against
`H2O2(0mM)` and uses the default Benjamini-Hochberg adjustment for p-values.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  tidy_dd,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "H2O2(0mM)"
)

knitr::kable(dt_stats, digits = 3)
```

## Plot doubling time comparisons

This plot shows mean doubling time with error bars and comparison brackets
against `H2O2(0mM)`.

```{r doubling-time-plot, fig.width=8, fig.height=5}
plot_doubling_time(
  tidy_dd,
  comparison_col = "condition",
  compare_to = "H2O2(0mM)",
  palette_name = "Dark2"
)
```

## Fit and plot one representative growth curve

This section fits a logistic model to the first sample in the reference
condition and overlays observed and fitted values.

```{r fit-one-sample}
sample_to_fit <- tidy_dd |>
  filter(condition == "H2O2(0mM)") |>
  distinct(sample) |>
  slice(1) |>
  pull(sample)

fit <- tidy_dd |>
  filter(sample == sample_to_fit) |>
  fit_growth_curve(model = "logistic")

extract_params(fit)
```

```{r plot-fit, fig.width=8, fig.height=5}
plot_fitted_curve(fit)
```
