---
title: "ScBS181 growkar workflow example"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "scbs181-growkar-workflow-files/figure-gfm/"
)

if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  pkg_root <- "."
} else {
  candidates <- c(".", "..", "../..")
  match_idx <- which(file.exists(file.path(candidates, "DESCRIPTION")))
  pkg_root <- if (length(match_idx) > 0L) candidates[[match_idx[[1]]]] else NULL
}

if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)

sc_colors <- c("blue4", "#E41A1C", "gold", "black")
```

## Import and validate data

This example reads `ScBS181_OD.txt`, converts it to the canonical tidy format
used by `growkar`, and validates the result.

```{r prepare-data}
sc_path <- if (file.exists("ScBS181_OD.txt")) {
  "ScBS181_OD.txt"
} else {
  system.file("extdata", "ScBS181_OD.txt", package = "growkar")
}

sc_raw <- read.delim(sc_path, check.names = FALSE)
tidy_sc <- as_tidy_growth_data(sc_raw)
validate_growth_data(tidy_sc)

head(tidy_sc)
```

## Create the canonical SummarizedExperiment

`growkar` can also package the processed data into a lightweight
`growkar_data` object and coerce it into a `SummarizedExperiment` for
Bioconductor-oriented workflows.

```{r summarized-experiment}
growkar_obj <- as_growkar(tidy_sc)
se <- methods::as(growkar_obj, "SummarizedExperiment")

selected_conditions <- c("Sc(100)", "Sc(50)", "Sc(25)", "Sc(0)")

keep <- SummarizedExperiment::colData(se)$condition %in% selected_conditions
se <- se[, keep]
selected_levels <- unique(as.character(SummarizedExperiment::colData(se)$condition))
SummarizedExperiment::colData(se)$condition <- factor(
  SummarizedExperiment::colData(se)$condition,
  levels = selected_levels
)

se
```

## Analyze using SummarizedExperiment metadata

The rolling-window doubling-time summary can also be computed directly on the
`SummarizedExperiment` object and retrieved from `metadata(se)$growth_metrics`.

```{r summarized-experiment-metrics}
se <- growth_metrics(
  se,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "Sc(0)"
)

se_metrics <- S4Vectors::metadata(se)$growth_metrics |>
  dplyr::mutate(condition = factor(.data$condition, levels = selected_conditions)) |>
  dplyr::arrange(.data$condition)

knitr::kable(se_metrics, digits = 3)
```

## Plot growth curves with averaged replicates

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  se,
  average_replicates = TRUE,
  colour_col = "condition",
  custom_colors = sc_colors
)
```

## Summarize doubling time from averaged replicates

This summary computes replicate-level doubling times for the selected Sc
conditions using the rolling-window method.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  se,
  method = "rolling_window",
  comparison_col = "condition",
  compare_to = "Sc(0)"
)

dt_stats <- dt_stats |>
  dplyr::mutate(condition = factor(.data$condition, levels = selected_conditions)) |>
  dplyr::arrange(.data$condition)

knitr::kable(dt_stats, digits = 3)
```

## Plot doubling time with error bars and p-value annotations

```{r doubling-time-plot, fig.width=8, fig.height=5}
plot_doubling_time(
  se,
  comparison_col = "condition",
  compare_to = "Sc(0)",
  average_replicates = FALSE,
  method = "rolling_window",
  custom_colors = sc_colors
)
```
