---
title: "ScBS181 growkar workflow example"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "scbs181-growkar-workflow-files/figure-gfm/"
)

if (requireNamespace("pkgload", quietly = TRUE) && file.exists("DESCRIPTION")) {
  pkg_root <- "."
} else {
  candidates <- c(".", "..", "../..")
  match_idx <- which(file.exists(file.path(candidates, "DESCRIPTION")))
  pkg_root <- if (length(match_idx) > 0L) candidates[[match_idx[[1]]]] else NULL
}

if (!is.null(pkg_root) && requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)

sc_colors <- c("blue4", "#E41A1C", "gold", "black")
```

## Prepare data

This example reads `ScBS181_OD.txt`, converts it to the canonical tidy format
used by `growkar`, and validates the result.

```{r prepare-data}
sc_path <- if (file.exists("ScBS181_OD.txt")) {
  "ScBS181_OD.txt"
} else {
  system.file("extdata", "ScBS181_OD.txt", package = "growkar")
}

sc_raw <- read.delim(sc_path, check.names = FALSE)
tidy_sc <- as_tidy_growth_data(sc_raw)
validate_growth_data(tidy_sc)

sc_levels <- tidy_sc |>
  dplyr::distinct(condition) |>
  dplyr::mutate(
    concentration = as.numeric(sub("^Sc\\(([-0-9.]+)\\)$", "\\1", .data$condition))
  ) |>
  dplyr::arrange(.data$concentration) |>
  dplyr::pull(condition)

tidy_sc <- dplyr::mutate(
  tidy_sc,
  condition = factor(.data$condition, levels = sc_levels)
)

selected_conditions <- c("Sc(100)", "Sc(50)", "Sc(25)", "Sc(0)")

tidy_sc <- tidy_sc |>
  dplyr::filter(.data$condition %in% selected_conditions) |>
  dplyr::mutate(condition = factor(.data$condition, levels = selected_conditions))

averaged_sc <- tidy_sc |>
  dplyr::group_by(.data$condition, .data$time) |>
  dplyr::summarise(od = mean(.data$od, na.rm = TRUE), .groups = "drop") |>
  dplyr::rename(sample = .data$condition) |>
  dplyr::mutate(sample = factor(.data$sample, levels = selected_conditions))

head(tidy_sc)
```

## Plot growth curves with averaged replicates

```{r average-growth-curve, fig.width=8, fig.height=5}
plot_growth_curve(
  tidy_sc,
  average_replicates = TRUE,
  colour_col = "sample",
  custom_colors = sc_colors
)
```

## Summarize doubling time from averaged replicates

This summary computes doubling time from the averaged growth trajectory for each
condition using the rolling-window method.

```{r doubling-time-summary}
dt_stats <- summarize_growth_metrics(
  tidy_sc,
  method = "rolling_window",
  average_replicates = TRUE
)

dt_stats <- dt_stats |>
  dplyr::mutate(sample = factor(.data$sample, levels = selected_conditions)) |>
  dplyr::arrange(.data$sample)

knitr::kable(dt_stats, digits = 3)
```

## Plot averaged doubling time

```{r doubling-time-plot, fig.width=8, fig.height=5}
ggplot2::ggplot(
  dt_stats,
  ggplot2::aes(x = .data$sample, y = .data$doubling_time, fill = .data$sample)
) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::scale_fill_manual(values = sc_colors) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "none", panel.grid.minor = ggplot2::element_blank()) +
  ggplot2::labs(x = "Condition", y = "Doubling time")
```

## Rolling-window exponential phase in all averaged samples

This section shows the highest-ranked rolling-window exponential interval for
each averaged Sc condition and the mean start and end times across all
conditions.

```{r detect-exponential-phase}
phase_tbl <- dplyr::bind_rows(
  lapply(split(averaged_sc, averaged_sc$sample), function(sample_data) {
    detect_exponential_phase(sample_data) |>
      dplyr::slice_head(n = 1)
  })
)

phase_tbl <- phase_tbl |>
  dplyr::arrange(.data$sample)

knitr::kable(phase_tbl, digits = 3)

phase_average <- phase_tbl |>
  dplyr::summarise(
    mean_start_time = mean(.data$start_time, na.rm = TRUE),
    mean_end_time = mean(.data$end_time, na.rm = TRUE)
  )

knitr::kable(phase_average, digits = 3)
```
