---
title: "growkar-introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{growkar-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(growkar)
library(dplyr)
library(knitr)
data(yeast_growth_data)
```

## Overview

`growkar` provides a tidy workflow for microbial growth curve analysis from
optical density time-series data. Version 2 centers on a canonical input format
with columns `sample`, `time`, and `od`, while preserving compatibility with
legacy wide plate-reader data.

The core workflow is:

1. Convert data into canonical tidy form.
2. Validate before downstream analysis.
3. Explore growth curves visually.
4. Compute empirical summaries and detect the exponential phase.
5. Fit logistic or Gompertz models where a full-curve summary is useful.

## Input formats and tidy conversion

```{r tidy-conversion}
tidy_growth <- as_tidy_growth_data(yeast_growth_data)

head(tidy_growth)
```

The canonical columns are `sample`, `time`, and `od`. Additional metadata such
as `condition` and `replicate` can be carried alongside them.

```{r validate-input}
validate_growth_data(tidy_growth)
```

## Exploratory plotting and quality control

Observed growth curves can be plotted directly from tidy data:

```{r plot-averaged-curves}
plot_growth_curve(
  tidy_growth,
  average_replicates = TRUE,
  colour_col = "condition",
  palette_name = "Dark2"
)
```

Replicate-level faceting remains available when averaging is disabled:

```{r plot-replicate-facets}
plot_growth_curve(
  tidy_growth,
  average_replicates = FALSE,
  colour_col = "condition",
  facet_col = "replicate",
  palette_name = "Dark2"
)
```

## Empirical summaries

```{r summarize-metrics}
metrics <- summarize_growth_metrics(
  tidy_growth,
  method = "rolling_window",
  average_replicates = TRUE
)

knitr::kable(metrics, digits = 3)
```

For a single sample, the empirical growth-rate estimate can be inspected in
more detail:

```{r single-sample-growth-rate}
sample_id <- unique(tidy_growth$sample)[1]

gr <- compute_growth_rate(
  filter(tidy_growth, sample == sample_id),
  method = "rolling_window"
)

knitr::kable(gr, digits = 3)
```

The doubling time helper is useful when you already have growth-rate values:

```{r doubling-time-helper}
knitr::kable(
  tibble(
    sample = gr$sample,
    growth_rate = gr$mu,
    doubling_time = compute_doubling_time(gr$mu)
  ),
  digits = 3
)
```

## Exponential phase detection

`detect_exponential_phase()` returns the ranked candidate windows and metadata
describing whether the chosen interval required any degraded fallback.

```{r detect-phase}
phase_tbl <- detect_exponential_phase(
  filter(tidy_growth, sample == sample_id)
)

knitr::kable(head(phase_tbl), digits = 3)
```

## Model fitting and diagnostics

For a full-curve model-based summary, fit one of the supported parametric
models:

```{r fit-logistic-model}
cg_fit <- tidy_growth |>
  filter(sample == sample_id) |>
  fit_growth_curve(model = "logistic")

extract_params(cg_fit)
summary(cg_fit)
```

Failed fits remain machine-readable and do not crash downstream helpers:

```{r failed-fit-example}
flat_fit <- fit_growth_curve(
  tibble(
    sample = "flat",
    time = 0:4,
    od = rep(0.2, 5)
  ),
  model = "logistic"
)

summary(flat_fit)
```

Observed and fitted values can be visualized together:

```{r plot-fitted-curve}
plot_fitted_curve(cg_fit)
```

## Migration from legacy growkar

Older workflows can continue to use:

- `calculate_growth_rate()`
- `calculate_growthrate_from_defined_time()`
- `calculate_growthrate_from_defined_logphase()`

These wrappers remain available, but they emit deprecation warnings and call
the tidy-v2 internals. New code should prefer:

- `compute_growth_rate()`
- `summarize_growth_metrics()`
- `detect_exponential_phase()`
- `fit_growth_curve()`

## Limitations and interpretation notes

- Growth-rate estimates rely on log-linear behavior and can be degraded for
  very sparse or flat curves.
- Zero OD values are allowed in the data, but log-based methods may warn and
  ignore them.
- A failed model fit still returns a `growkar_fit` object with status and
  diagnostics rather than throwing a cryptic `nls` error.

## Session information

```{r session-information}
sessionInfo()
```
