---
title: "growkar-introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{growkar-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
pkg_root <- if (file.exists("DESCRIPTION")) "." else ".."

if (requireNamespace("pkgload", quietly = TRUE) &&
    file.exists(file.path(pkg_root, "DESCRIPTION"))) {
  pkgload::load_all(pkg_root, export_all = FALSE, helpers = FALSE, quiet = TRUE)
}

library(growkar)
library(dplyr)
library(knitr)
data(yeast_growth_data)
```

## Background

High-throughput microbial growth assays are widely used to characterize
phenotypic responses to genetic perturbations, environmental conditions, and
drug treatments. These assays generate time-resolved measurements across large
numbers of samples, analogous to other high-throughput experimental platforms in
functional genomics.

While tools exist for transcriptomic and epigenomic data analysis within the
Bioconductor ecosystem, there is a relative lack of standardized infrastructure
for analyzing growth-based phenotypic data and integrating it with omics
datasets.

The `growkar` package addresses this gap by providing a scalable and
reproducible framework for high-throughput microbial phenotyping from growth
assays. It uses Bioconductor data structures to represent assay measurements
and derived phenotypes, while still accepting tidy and wide inputs as import
adapters for exploratory analysis and visualization.

The primary workflow is:

1. Import assay data and build a standardized `SummarizedExperiment`.
2. Validate the SE structure before downstream analysis.
3. Explore growth curves visually.
4. Compute empirical summaries and detect the exponential phase.
5. Fit logistic or Gompertz models where a full-curve summary is useful.

## Bioconductor container workflow

For Bioconductor-oriented workflows, growth measurements can be represented as a
`SummarizedExperiment`, with OD values stored in the assay, time values in
`rowData()`, sample annotations in `colData()`, and derived summaries in
`metadata()`.

```{r summarized-experiment}
growkar_obj <- as_growkar(yeast_growth_data)
se <- methods::as(growkar_obj, "SummarizedExperiment")
se <- growth_metrics(se, method = "rolling_window", average_replicates = TRUE)

S4Vectors::metadata(se)$growth_metrics
```

This workflow is useful when growth phenotypes need to be carried forward into
other Bioconductor analyses or linked to omics-derived sample annotations.

The canonical SE layout in `growkar` is:

- `assay(se, "od")`: OD matrix with timepoints in rows and samples in columns
- `rowData(se)`: timepoint metadata
- `colData(se)`: sample metadata
- `metadata(se)`: derived results such as growth metrics, phase windows, and fits

Helper accessors make these components easier to inspect:

```{r se-accessors}
growth_assay(se)[1:3, 1:3]
timepoints(se)
sample_data(se)
```

## Input formats and tidy conversion

```{r tidy-conversion}
tidy_growth <- as_tidy_growth_data(yeast_growth_data)

head(tidy_growth)
```

The canonical columns are `sample`, `time`, and `od`. Additional metadata such
as `condition` and `replicate` can be carried alongside them.

```{r validate-input}
validate_growth_data(tidy_growth)
```

The same tidy data can also be converted back into a `SummarizedExperiment`
whenever a container-based workflow is needed.

```{r summarized-experiment-direct}
as_summarized_experiment(yeast_growth_data)
```

## Exploratory plotting and quality control

Observed growth curves can be plotted directly from tidy data:

```{r plot-averaged-curves}
plot_growth_curve(
  tidy_growth,
  average_replicates = TRUE,
  colour_col = "condition",
  palette_name = "Dark2"
)
```

Replicate-level faceting remains available when averaging is disabled:

```{r plot-replicate-facets}
plot_growth_curve(
  tidy_growth,
  average_replicates = FALSE,
  colour_col = "condition",
  facet_col = "replicate",
  palette_name = "Dark2"
)
```

## Empirical summaries

```{r summarize-metrics}
metrics <- summarize_growth_metrics(
  tidy_growth,
  method = "rolling_window",
  average_replicates = TRUE
)

knitr::kable(metrics, digits = 3)
```

For a single sample, the empirical growth-rate estimate can be inspected in
more detail:

```{r single-sample-growth-rate}
sample_id <- unique(tidy_growth$sample)[1]

gr <- compute_growth_rate(
  filter(tidy_growth, sample == sample_id),
  method = "rolling_window"
)

knitr::kable(gr, digits = 3)
```

The doubling time helper is useful when you already have growth-rate values:

```{r doubling-time-helper}
knitr::kable(
  tibble(
    sample = gr$sample,
    growth_rate = gr$mu,
    doubling_time = compute_doubling_time(gr$mu)
  ),
  digits = 3
)
```

## Exponential phase detection

`detect_exponential_phase()` returns the ranked candidate windows and metadata
describing whether the chosen interval required any degraded fallback.

```{r detect-phase}
phase_tbl <- detect_exponential_phase(
  filter(tidy_growth, sample == sample_id)
)

knitr::kable(head(phase_tbl), digits = 3)
```

## Model fitting and diagnostics

For a full-curve model-based summary, fit one of the supported parametric
models:

```{r fit-logistic-model}
cg_fit <- tidy_growth |>
  filter(sample == sample_id) |>
  fit_growth_curve(model = "logistic")

extract_params(cg_fit)
summary(cg_fit)
```

Failed fits remain machine-readable and do not crash downstream helpers:

```{r failed-fit-example}
flat_fit <- fit_growth_curve(
  tibble(
    sample = "flat",
    time = 0:4,
    od = rep(0.2, 5)
  ),
  model = "logistic"
)

summary(flat_fit)
```

Observed and fitted values can be visualized together:

```{r plot-fitted-curve}
plot_fitted_curve(cg_fit)
```

## Supported API

The supported SE-native interface includes:

- `as_tidy_growth_data()`
- `validate_growth_data()`
- `compute_growth_rate()`
- `summarize_growth_metrics()`
- `detect_exponential_phase()`
- `fit_growth_curve()`

## Limitations and interpretation notes

- Growth-rate estimates rely on log-linear behavior and can be degraded for
  very sparse or flat curves.
- Zero OD values are allowed in the data, but log-based methods may warn and
  ignore them.
- A failed model fit still returns a `growkar_fit` object with status and
  diagnostics rather than throwing a cryptic `nls` error.

## Session information

```{r session-information}
sessionInfo()
```
