---
title: "growkar-introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{growkar-introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(growkar)
library(ggplot2)
```

## Overview

`growkar` provides a tidy workflow for microbial growth curve analysis from
optical density time-series data. Version 2 centers on a canonical input format
with columns `sample`, `time`, and `od`, while preserving compatibility with
legacy wide plate-reader data.

## Canonical tidy data

```{r}
tidy_growth <- as_tidy_growth_data(yeast_growth_data)

tidy_growth
```

## Plot observed curves

```{r}
plot_growth_curve(
  tidy_growth,
  average_replicates = TRUE,
  colour_col = "condition"
)
```

## Summarize empirical growth metrics

```{r}
metrics <- summarize_growth_metrics(
  tidy_growth,
  method = "rolling_window",
  average_replicates = TRUE
)

metrics
```

## Fit a parametric growth model

```{r}
cg_fit <- tidy_growth |>
  dplyr::filter(sample == "Cg_R1") |>
  fit_growth_curve(model = "logistic")

extract_params(cg_fit)
plot_fitted_curve(cg_fit)
```
